Causal Compatibility Inequalities Admitting of Quantum Violations in the Triangle structure

### 2017-06-01 documents/TriangleScenarioArXiv/ts.tex
Improved the discussion of d-seperation

    Now that the notion of a causal structure $\graph$ is defined along with some standard graph-theoretic notation, it is now possible to precisely define the notion of compatibility between $\graph$ and a probability distribution $\prob[\nodes]$ defined over the nodes of $\graph$. Intuitively, the absence of certain causal connections within $\graph$ should limit the set of possible observations that can be made on $\graph$. Specifically, the causal structure $\graph$ hypothesizes that if each variable $n \in \nodes$ is causally influenced by its parents $\Pa[\graph]{n}$, then each $n$ should be conditionally independent of all other variables:
    \[ \prob[\nodes] = \prod_{n \in \nodes} \prob[n \mid \Pa[\graph]{n}] = \prob[n_1 \mid \Pa[\graph]{n_1}] \times \cdots \times \prob[n_k \mid \Pa[\graph]{n_k}] \eq \label{eq:compatibility_as_factorization}\]
    % , except its own descendants when conditioned over $\Pa[\graph]{n}$. This is known as the \term{local Markov property}:\footnote{Here $A \indep B \mid C$ denotes statistical independence between $A$ and $B$ conditional on $C$ where $A,B,C \subseteq \nodes$, i.e. for a given probability distribution $\prob[\nodes]$, $\prob[AB \mid C] = \prob[A \mid C]\prob[B \mid C]$.}
    % \[ n \indep \br{\nodes \setminus \De[\graph]{n}} \mid \Pa[\graph]{n} \quad \forall n \in \nodes \eq \label{eq:local_markov_property} \]
    If a given distribution $\prob[\nodes]$ defined over \textit{all} of the nodes $\nodes$ of $\graph$ satisfies \cref{eq:compatibility_as_factorization}, then $\prob[\nodes]$ is said to be \term{compatible with $\graph$}. The conditional distributions in \cref{eq:compatibility_as_factorization} (i.e. $\{\prob[n \mid \Pa[\graph]{n}] \mid n \in \nodes\}$) are referred to as a set of \term{causal parameters for $\graph$}. If a distribution $\prob[\nodes]$ cannot be factorized according to \cref{eq:compatibility_as_factorization}, $\prob[\nodes]$ is said to be \term{incompatible} with $\graph$. When one is specified with a joint distribution $\prob[\nodes]$ defined over all nodes of a causal structure $\graph$, it is trivially easy to verify whether or not $\prob[\nodes]$ is compatibility with $\graph$; \cref{sec:satisfaction_of_d_sep_relations} describes in detail how this is accomplished.

    A challenge is present when one is supplied with a partial observation, i.e. a joint distribution $\prob[\nodes_{O}]$ where $\nodes_{O} \subset \nodes$ is some subset of variables referred to as the \term{observable nodes} $\nodes_O$. In such cases, \cref{eq:compatibility_as_factorization} can not be verified by direction calculation because it is certainly possible that the parents of some observable node are themselves unobservable $\Pa[\graph]{\nodes_{O}} \not \subseteq \nodes_{O}$. In such cases, compatibility between $\graph$ and $\prob[\nodes_{O}]$ depends on the existence or non-existence of a set of causal parameters for $\graph$ such that $\prob[\nodes_{O}] = \prod_{n \in \nodes_{O}} \prob[n \mid \Pa[\graph]{n}]$. The unobservable nodes are termed \term{latent nodes} $\nodes_L = \nodes \setminus \nodes_{O}$ representing hidden random variables that are hidden either by some fundamental process or cannot be measured due to other limitations.
    \todo[TC]{Explain why this is difficult more.}
    \todo[TC]{Figure out if I should include d-sep here}
    \todo[TC]{Explain what this work and inflation accomplishes regarding this}

    In practice however, checking causal parameters systemically is intractable, especially when the cardinality (number of outcomes) of the latent variables is unspecified. Instead, \term{causal compatibility inequalities}\footnote{We refer to these inequalities as ``causal compatibility inequalities'' instead of simply Bell inequalities for two reasons. Firstly, ``Bell inequalities'' usually are associated specifically with the Bell structure. Secondly, the inequalities derived in this work are fundamentally distinct from a typical Bell inequality in that these inequalities are \textit{polynomial} over $\prob[\nodes_{O}]$ instead of \textit{linear}.} are probabilistic inequalities defined over $\prob[\nodes_{O}]$ and its induced marginals which constraint the set of compatible probability distributions. If a probability distribution $\prob[\nodes_{O}]$ happens to \textit{violate} any causal compatibility inequality, then that distribution deemed \textit{incompatible}. In \cref{sec:deriving_inequalities}, a brief discussion of existing techniques for deriving these inequalities will take place.

@@

    The remaining inequalities in \cref{sec:found_inequalities} were derived using the Web inflation of \cref{fig:the_web_inflation}. The maximal pre-injectable sets of the Wagon-Wheel inflation along with their ancestral dependences can be found in \cref{eq:pre_injectable_triangle_structure_web}.
    \begin{equation*}
        \eq \label{eq:pre_injectable_triangle_structure_web}
        \begin{gathered}
            \textbf{Maximal Pre-injectable Sets} \\
            \bc{A_1, B_1, C_1, A_4, B_4, C_4} \\
            \bc{A_1, B_2, C_3, A_4, B_3, C_2} \\
            \bc{A_2, B_3, C_1, A_3, B_2, C_4} \\
            \bc{A_2, B_4, C_3, A_3, B_1, C_2} \\
            \bc{A_1, B_3, C_4} \\
            \bc{A_1, B_4, C_2} \\
            \bc{A_2, B_1, C_4} \\
            \bc{A_2, B_2, C_2} \\
            \bc{A_3, B_3, C_3} \\
            \bc{A_3, B_4, C_1} \\
            \bc{A_4, B_1, C_3} \\
            \bc{A_4, B_2, C_1}
        \end{gathered}
        \qquad
        \begin{gathered}
            \textbf{Ancestral Independences} \\
            \bc{A_1, B_1, C_1} \ancestralindep \bc{A_4, B_4, C_4} \\
            \bc{A_1, B_2, C_3} \ancestralindep \bc{A_4, B_3, C_2} \\
            \bc{A_2, B_3, C_1} \ancestralindep \bc{A_3, B_2, C_4} \\
            \bc{A_2, B_4, C_3} \ancestralindep \bc{A_3, B_1, C_2} \\
            \bc{A_1} \ancestralindep \bc{B_3} \ancestralindep \bc{C_4} \\
            \bc{A_1} \ancestralindep \bc{B_4} \ancestralindep \bc{C_2} \\
            \bc{A_2} \ancestralindep \bc{B_1} \ancestralindep \bc{C_4} \\
            \bc{A_2} \ancestralindep \bc{B_2} \ancestralindep \bc{C_2} \\
            \bc{A_3} \ancestralindep \bc{B_3} \ancestralindep \bc{C_3} \\
            \bc{A_3} \ancestralindep \bc{B_4} \ancestralindep \bc{C_1} \\
            \bc{A_4} \ancestralindep \bc{B_1} \ancestralindep \bc{C_3} \\
            \bc{A_4} \ancestralindep \bc{B_2} \ancestralindep \bc{C_1}
        \end{gathered}
    \end{equation*}
    Analogously to the Wagon-Wheel inflation, these pre-injectable sets form a marginal scenario $\mscenario$ which defines an incidence matrix $M$ that has $\abs{\Events{\mscenario}} = 4 \cdot 4^{6} + 8 \cdot 4^3 = 16,896$ rows, $\abs{\Events{\jointvar}} = 4^{12} = 16,777,216$ and $201,326,592$ non-zero entires.

>>>>

    Now that the notion of a causal structure $\graph$ is defined along with some standard graph-theoretic notation, it is now possible to precisely define the notion of compatibility between $\graph$ and a probability distribution $\prob[\nodes]$ defined over the nodes of $\graph$. Intuitively, the absence of certain causal connections within $\graph$ should limit the set of possible observations that can be made on $\graph$. Specifically, the causal structure $\graph$ hypothesizes that if each variable $n \in \nodes$ is causally influenced by its parents $\Pa[\graph]{n}$, then each $n$ should be conditionally independent of all other variables:
    \[ \prob[\nodes] = \prod_{n \in \nodes} \prob[n \mid \Pa[\graph]{n}] = \prob[n_1 \mid \Pa[\graph]{n_1}] \times \cdots \times \prob[n_k \mid \Pa[\graph]{n_k}] \eq \label{eq:compatibility_as_factorization}\]
    % , except its own descendants when conditioned over $\Pa[\graph]{n}$. This is known as the \term{local Markov property}:\footnote{Here $A \indep B \mid C$ denotes statistical independence between $A$ and $B$ conditional on $C$ where $A,B,C \subseteq \nodes$, i.e. for a given probability distribution $\prob[\nodes]$, $\prob[AB \mid C] = \prob[A \mid C]\prob[B \mid C]$.}
    % \[ n \indep \br{\nodes \setminus \De[\graph]{n}} \mid \Pa[\graph]{n} \quad \forall n \in \nodes \eq \label{eq:local_markov_property} \]
    If a given distribution $\prob[\nodes]$ defined over \textit{all} of the nodes $\nodes$ of $\graph$ satisfies \cref{eq:compatibility_as_factorization}, then $\prob[\nodes]$ is said to be \term{compatible with $\graph$}. The conditional distributions in \cref{eq:compatibility_as_factorization} (i.e. $\{\prob[n \mid \Pa[\graph]{n}] \mid n \in \nodes\}$) are referred to as a set of \term{causal parameters for $\graph$}. If a distribution $\prob[\nodes]$ cannot be factorized according to \cref{eq:compatibility_as_factorization}, $\prob[\nodes]$ is said to be \term{incompatible} with $\graph$. The compatibility of a distribution $\prob[\nodes]$ with a causal structure $\graph$ pursuant to \Cref{eq:compatibility_as_factorization} is equivalent to the requirement that $\prob[\nodes]$ obeys all of the conditional independence constraints implied by the \textit{d-separation relations} between variables/nodes in $\graph$~\cite{Pearl_2009}. When one is specified with a joint distribution $\prob[\nodes]$ defined over all nodes of a causal structure $\graph$, it is trivially easy to verify whether or not $\prob[\nodes]$ is compatible with $\graph$ by simply computing the causal parameters induced by $\prob[\nodes]$ and checking the equality of \cref{eq:compatibility_as_factorization}.

    A challenge is presented when one is supplied with a partial observation, i.e. a joint distribution $\prob[\nodes_{O}]$ where $\nodes_{O} \subset \nodes$ is some subset of variables referred to as the \term{observable nodes} $\nodes_O$. In such cases, $\prob[\nodes_{O}]$ does not induce a unique set of causal parameters for $\graph$ and \cref{eq:compatibility_as_factorization} can not be verified by direction calculation. Instead, compatibility between $\graph$ and $\prob[\nodes_{O}]$ depends on the existence or non-existence of a set of causal parameters for $\graph$ such that $\prob[\nodes_{O}] = \prod_{n \in \nodes_{O}} \prob[n \mid \Pa[\graph]{n}]$. The complementary, unobservable nodes are termed \term{latent nodes} $\nodes_L = \nodes \setminus \nodes_{O}$ representing hidden random variables that are hidden either by some fundamental process or cannot be measured due to other limitations.

    There are several approaches to tackling the compatibility problem when dealing with latent variables; there are two naive approaches worth mentioning here. The first is to recognize that the conditional independence constraints implied by d-separation relations over the \textit{observable} variables form a family of necessary constraints for $\prob[\nodes_{O}]$. The failure to satisfy one of these observable independence relations immediately qualifies $\prob[\nodes_{O}]$ as incompatible with $\graph$. Often, the observable d-separation relations implied by a causal structure form an \textit{insufficient} set of constraints for compatible distributions. Such causal structures have been termed \textit{interesting}~\cite{Henson_2014}, and more involved classification techniques must be used~\cite{Inflation}. The principle focus of this paper is the Triangle structure and has been shown to be one of the smallest interesting causal structures~\cite{Henson_2014}. An alternative approach requires the cardinality (number of outcomes) of the latent variables to be fixed. In most realistic scenarios, the nature of the latent variables is completely unspecified. More powerful approaches find constraints for $\prob[\nodes]$ in the form of probabilistic inequalities defined over $\prob[\nodes_{O}]$ and its marginals which constraint the set of compatible probability distributions. We term these constraints \term{causal compatibility inequalities}\footnote{We refer to these inequalities as ``causal compatibility inequalities'' instead of simply Bell inequalities for two reasons. Firstly, ``Bell inequalities'' usually are associated specifically with the Bell structure. Secondly, the inequalities derived in this work are fundamentally distinct from a typical Bell inequality in that these inequalities are \textit{polynomial} over $\prob[\nodes_{O}]$ instead of \textit{linear}.} If a probability distribution $\prob[\nodes_{O}]$ happens to \textit{violate} any causal compatibility inequality, then that distribution deemed \textit{incompatible}. In \cref{sec:deriving_inequalities}, a brief discussion of existing techniques for deriving these inequalities will take place.

@@

    The remaining inequalities in \cref{sec:found_inequalities} were derived using the Web inflation of \cref{fig:the_web_inflation}. The maximal pre-injectable sets of the Wagon-Wheel inflation along with their ancestral dependences can be found in \cref{eq:pre_injectable_triangle_structure_web}.
    \begin{equation*}
        \eq \label{eq:pre_injectable_triangle_structure_web}
        \begin{gathered}
            \textbf{Maximal Pre-injectable Sets} \\
            \bc{A_1, B_1, C_1, A_4, B_4, C_4} \\
            \bc{A_1, B_2, C_3, A_4, B_3, C_2} \\
            \bc{A_2, B_3, C_1, A_3, B_2, C_4} \\
            \bc{A_2, B_4, C_3, A_3, B_1, C_2} \\
            \bc{A_1, B_3, C_4} \\
            \bc{A_1, B_4, C_2} \\
            \bc{A_2, B_1, C_4} \\
            \bc{A_2, B_2, C_2} \\
            \bc{A_3, B_3, C_3} \\
            \bc{A_3, B_4, C_1} \\
            \bc{A_4, B_1, C_3} \\
            \bc{A_4, B_2, C_1}
        \end{gathered}
        \qquad
        \begin{gathered}
            \textbf{Ancestral Independences} \\
            \bc{A_1, B_1, C_1} \ancestralindep \bc{A_4, B_4, C_4} \\
            \bc{A_1, B_2, C_3} \ancestralindep \bc{A_4, B_3, C_2} \\
            \bc{A_2, B_3, C_1} \ancestralindep \bc{A_3, B_2, C_4} \\
            \bc{A_2, B_4, C_3} \ancestralindep \bc{A_3, B_1, C_2} \\
            \bc{A_1} \ancestralindep \bc{B_3} \ancestralindep \bc{C_4} \\
            \bc{A_1} \ancestralindep \bc{B_4} \ancestralindep \bc{C_2} \\
            \bc{A_2} \ancestralindep \bc{B_1} \ancestralindep \bc{C_4} \\
            \bc{A_2} \ancestralindep \bc{B_2} \ancestralindep \bc{C_2} \\
            \bc{A_3} \ancestralindep \bc{B_3} \ancestralindep \bc{C_3} \\
            \bc{A_3} \ancestralindep \bc{B_4} \ancestralindep \bc{C_1} \\
            \bc{A_4} \ancestralindep \bc{B_1} \ancestralindep \bc{C_3} \\
            \bc{A_4} \ancestralindep \bc{B_2} \ancestralindep \bc{C_1}
        \end{gathered}
    \end{equation*}
    Analogously to the Wagon-Wheel inflation, these pre-injectable sets form a marginal scenario $\mscenario$ which defines an incidence matrix $M$ that has $\abs{\Events{\mscenario}} = 4 \cdot 4^{6} + 8 \cdot 4^3 = 16,896$ rows, $\abs{\Events{\jointvar}} = 4^{12} = 16,777,216$ columns and $201,326,592$ non-zero entires.
